---
title: "repdata-012 PA1: Activity Monitoring Data "
author: "zenelg"
date: "Wednesday, March 04, 2015"
output: html_document
---

**LOADING AND PREPROCESSING THE DATA**
```{r echo=TRUE}
library(dplyr)
library(tidyr)
library(lattice)

activity <- read.csv("activity.csv", header=TRUE, stringsAsFactors=FALSE)
activity$date <- as.POSIXct(activity$date, format="%Y-%m-%d")
activity$dayofweek <- weekdays(activity$date, abbr = TRUE)
```

**WHAT IS MEAN TOTAL NUMBER OF STEPS TAKEN PER DAY?**

    What is mean total number of steps taken per day? (ignoring the missing values in the dataset)
    
```{r echo=TRUE}
# Aggregates steps by date
aggdata <-aggregate(steps ~ date, data = activity, FUN = mean)
#activityCC <-  activity[complete.cases(activity),]
#aggdatamed <-aggregate(steps ~ date, data = activityCC, FUN = median, na.rm=TRUE)
# Hist mean steps by day
hist(aggdata$steps, main="mean steps by day")
# Calculates the mean of steps of the dataset
meanstepsNA <- mean(na.omit(aggdata$steps))
#aggdata <-aggregate(steps ~ date, data = activity, FUN = median)
medianstepsNA <- median(na.omit(aggdata$steps))

meanstepsNA
medianstepsNA  #Review

```

    *`r meanstepsNA` mean total number of steps.*
    
**WHAT IS MEAN TOTAL NUMBER OF STEPS TAKEN PER DAY?**

    Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
  
```{r echo=TRUE}
# Grouping by interval and summarizing mean by dates for plot (na removed)
by_interval <- group_by(activity, interval)
by_intervalmean <- summarize(by_interval, mean_steps = mean(steps, na.rm=TRUE))
#intervals <- strptime(sprintf("%04d", as.numeric(names(meanSteps))), format="%H%M")
with(by_intervalmean, plot(interval, mean_steps, type="l"))

```

    *0835 interval contains the maximum number of steps on average across all the days in the dataset.*

**IMPUTING MISSING VALUES**

    Calculate and report the total number of missing values in the dataset 
    
```{r echo=TRUE}
missingvalues <- nrow(activity) - sum(complete.cases(activity))

```

    *The total number of missing values in the dataset is `r missingvalues`*
    
    *We will use the mean per interval to fill the missing values in the dataset 
in a new dataset activity2*
    
```{r echo=TRUE}
activity2 <- select(activity, steps, date, interval, dayofweek)

for(n in 1:nrow(activity2)) {
    
    index <- activity2[n,3]
    temp <- by_intervalmean[by_intervalmean$interval == index, ]
    if (is.na(activity2[n,1])) activity2[n,1] <- temp$mean_steps else flag <-                 0
    
}

```

    Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
    
```{r echo=TRUE}
par(mfrow = c(2,1))
# With NAs
hist(aggdata$steps, main="mean steps by day - NA")
# NAs estimated as mean of Day
    # Aggregates steps by date
aggdata <-aggregate(steps ~ date, data = activity2, FUN = mean)
    # Hist mean steps by day
hist(aggdata$steps, main="mean steps by day")
# Calculates the mean of steps of the dataset
meansteps <- mean(na.omit(aggdata$steps))
meansteps

```

    Do these values differ from the estimates from the first part of the assignment? - *Yes.*
    
    What is the impact of imputing missing data on the estimates of the total daily number of steps? - *There is a difference in means with `r meansteps` mean total number of steps (NAs estimated by mean) vs `r meanstepsNA` mean total number of steps.*

**ARE THERE DIFFERENCES IN ACTIVITY PATTERNS BETWEEN WEEKDAYS AND WEEKENDS?**

    1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
    
```{r echo=TRUE}
#activity2 <- select(activity, steps, date, interval, dayofweek)
activity2[, "WKDorWND"] <- "weekday"
WKN <- activity2[ activity2$dayofweek %in% c("Sat","Sun"), ]
WKN[, "WKDorWND"] <- "weekend"
WKD <- activity2[ ! activity2$dayofweek %in% c("Sat","Sun"), ]
activity2 <- rbind(WKN,WKD)
activity2 <- arrange(activity2, date)

head(activity2)

```

    *See above for added column 'WKDorWND' for weekend or weekdays.*

    2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r echo=TRUE}
# Grouping by interval and summarizing mean by dates for plot (na removed)
activity2WD <- subset(activity2, WKDorWND=="weekday", select=steps:WKDorWND) 
by_interval2WD <- group_by(activity2WD, interval)
by_intervalmean2WD <- summarise(by_interval2WD, mean_steps = mean(steps, na.rm=TRUE))
by_intervalmean2WD$WKDorWND <- "weekday"

activity2WN <- subset(activity2, WKDorWND=="weekend", select=steps:WKDorWND) 
by_interval2WN <- group_by(activity2WN, interval)
by_intervalmean2WN <- summarise(by_interval2WN, mean_steps = mean(steps, na.rm=TRUE))
by_intervalmean2WN$WKDorWND <- "weekend"

by_intervalmean2b <- rbind(by_intervalmean2WD, by_intervalmean2WN)

xyplot(mean_steps ~ interval | WKDorWND, data = by_intervalmean2b, type='l',layout = c(1, 2))   

```

